Learning Constraints and Optimization Criteria

### 2015-10-30 Paper.tex
Implementing Angelika's suggestions and rewriting, part 1

\graphicspath{{Graphics/}}

@@

\newtheorem{definition}{Definition}[section]
\newtheorem{question}{Question}
\AfterEndEnvironment{definition}{\noindent\ignorespaces}

\theoremstyle{definition}
\newtheorem{example}{Example}
\newtheorem{experiment}{Experiment}[question]
\newtheorem{observation}{Observation}[question]

\newcommand{\anton}[1]{{\color{green}(#1)}}
\newcommand{\sam}[1]{{\color{red}(#1)}}

@@

\begin{abstract}
While there exist several approaches in the constraint programming community to learn a constraint theory, few of them have considered the learning of constraint optimization problems.
To alleviate this situation, we introduce an initial approach to learning first-order weighted MAX-SAT theories. 
It employs inductive logic programming techniques to learn a set of first-order clauses and then uses preference learning techniques to learn the weights of the clauses.
In order to learn these weighted clauses, the clausal optimization system uses examples of possible worlds and a set of preferences that state which examples are preferred over other ones.
The technique is also empirically evaluated on a number of examples.

\end{abstract}

@@

\section{Introduction}
An important class of declarative models is concerned with (discrete) constrained optimization problems.
These combine constraint satisfaction problems with an optimization function that specifies which solutions are optimal.
While these models are powerful and elegant, they are often also hard to obtain, certainly for non-experts \cite{Wallace:PrinciplesCP}.
This has motivated several researchers to investigate whether such models can be learned.

Learning constraints has been addressed in constraint programming (e.g. \cite{Beldiceanu:ModelSeeker}, \cite{bessiere2013constraint}) and inductive logic programming (e.g. \cite{DeRaedt:ClausalDiscovery}).
Recent efforts \cite{Lallouet:LearningCP} have attempted to combine these two fields.
Furthermore, \cite{campigotto2011active}) have learned propositional weighted MAX-SAT theories from examples.
Also, work in statistical relational learning (SRL) on learning the structure of Markov Logic networks (MLN) or other SRL models \cite{kok2005learning} is relevant, as MAP inference can be used to query for the most likely state.

@@

\paragraph{Clauses}
We use (function free) first-order clauses; which we now introduce.
An atom $p(t_1, t_2, ..., t_n)$ consists of a predicate symbol~$p$ that is applied to $n$ terms~$t_i$.
Terms are either constants or variables.
In this paper constants will be written in uppercase ($\mathit{John}$) and variables in lowercase ($\mathit{x}$).
A literal can be an atom $a$ (positive literal) or a negated atom $\lnot a$ (negative literal).
Clauses are disjunctions of literals $l_1 \lor l_2 \lor ... \lor l_k$ and are assumed to be universally quantified.
A clause can be expressed using a body and a head $\mathit{head} \leftarrow \mathit{body}$.
Hereby, positive literals are grouped in the head and negative literals are grouped in the body.
Sometimes clauses are also written as sets of literals $\{l_1, l_2, ..., l_k\}$.

@@

\paragraph{Optimization}
The Boolean Satisfiability Problem (SAT) attempts to determine whether a propositional formula can be satisfied.
Formulas can be rewritten into Conjunctive Normal Form (CNF), which consists of a conjunction of clauses (disjunctions).
An extension of this problem is the Maximum Satisfiability Problem (MAX-SAT), which is an optimization problem and attempts to maximize the number of satisfied clauses.
By assigning positive integer weights to clauses and maximizing the sum of the weights of satisfied clauses, one obtains the weighted MAX-SAT problem.

In related work \cite{campigotto2011active} weighted MAX-SAT problems are induced by learning propositional clauses and weights using examples and absolute scores.
\sam{Change this section}
Inspired by their results, our research tries to learn first-order logic clauses and weights using examples and rankings.
The resulting optimization problem can be seen as an extension of weighted MAX-SAT in which the clauses are in first-order logic.

\paragraph{Learning Constraints}
The goal of constraint learning is to automatically identify constraints in one or multiple examples.
This task is difficult because the search space of constraints is typically very large and, unlike in typical machine learning problems, there are usually only few examples to learn from.

@@

\section{Problem Statement}
\label{sec:problem}
We want to learn constraints and optimization criteria based on positive examples and user preferences.
The goal of this task is to simplify the process of generating formal representations, therefore, any input should be reasonably easy for a user to provide.

\subsection{Constraint Learning}
\emph{Given a set of examples and a limit $t$, the goal is to find maximally specific clauses that are satisfied by at least $t$ of the given examples.}
\\\\
Examples are Herbrand interpretations.
They consist of set of constants, the domain, and all ground atoms over these constants that are true.
The learned clauses are domain-independent and contain only variables.

\subsection{Learning Optimization Criteria}
\emph{Given a set of examples \sym{E} and a set of preferences \sym{P} over these examples, the goal is to learn soft constraints and weights such that the order described by these optimization criteria maximally corresponds with~\sym{P}.}
\\\\
Every preference $p \in \sym{P}$ describes a relative order $e_1 \succ e_2 \succ ... \succ e_n$ over a subset of examples $\{e_1, e_2, ..., e_n\} \subseteq \sym{E}$.
A preference $e_1 \succ e_2$ indicates that out of examples $e_1$ and $e_2$ the user prefers $e_1$.

The soft constraints consist of clauses \sym{C} that are satisfied by some but not all of the examples.
They can be identified using a system that implements the constraint learning task.
The weighted clauses \sym{W} are tuples $\mathbb{R} \times \sym{C}$ that can be used to calculate a score for any example $e$:
\begin{eqnarray}
  \label{eqn:score}
  score(e, \sym{W}) = \sum\limits_{(\mathit{w}, \mathit{c}) \in \sym{W}} \mathit{w} \cdot \mathbf{1}_{e \models c}
\end{eqnarray}

@@

  Assume the following first-order weighted MAX-SAT model \sym{M} describes the optimization problem to solve:
  \begin{flalign*}
    (w_1, c_1) &= (0.50, \mathit{low\_crime}(a) \leftarrow \mathit{live\_in}(a)) \\
    (w_2, c_2) &= (0.25, \mathit{school\_in}(a) \leftarrow \mathit{work\_in}(a)) \\
    (w_3, c_3) &= (1.00, \mathit{low\_crime}(a) \leftarrow \mathit{school\_in}(a)) \\
    (w_4, c_4) &= (-1.00, \mathit{false} \leftarrow \mathit{live\_in}(a) \land \mathit{cheap}(a))
  \end{flalign*}
  Let examples $e_1$, $e_2$ and~$e_3$ be interpretations that reason over the same city.
  Specifically, let all examples share constants (areas) $A_1$, $A_2$ and $A_3$ and ground atoms $\mathit{cheap}(A_1)$, $\mathit{cheap}(A_3)$, $\mathit{low\_crime}(A_1)$ and $\mathit{low\_crime}(A_2)$. Additionally, they specify:
  \begin{flalign*}
    &e_1: \mathit{live\_in}(A_1), \mathit{work\_in}(A_2), \mathit{school\_in(A_2)} \\
    &e_2: \mathit{live\_in}(A_2), \mathit{work\_in}(A_2), \mathit{school\_in(A_3)} \\
    &e_3: \mathit{live\_in}(A_2), \mathit{work\_in}(A_2), \mathit{school\_in(A_1)}
  \end{flalign*}

  The score of an example for the model~\sym{M} can be calculated using (\ref{eqn:score}).
  Example~$e_1$ is covered by clauses $c_1$, $c_2$ and~$c_3$, therefore, $score(e_1, \sym{M}) = 0.5 + 0.25 + 1 + 0 = 1.75$.
  Similarly, $score(e_2, \sym{M}) = -1$ and $score(e_3, \sym{M}) = 0$ can be computer.
  Based on these scores, the model \sym{M} would prefer $e_1$ over~$e_2$ ($e_1 \succ e_2$), $e_1$ over~$e_3$ ($e_1 \succ e_3$) and $e_2$ over~$e_3$ ($e_2 \succ e_3$).
  The aim of our research is to learn a model like \sym{M}, given such examples and preferences.
\end{example}

Apart from learning optimization criteria, this paper presents an approach to use these optimization criteria to find an optimal solution in practice.

@@

\section{Approach}
In this work, we use a two-step approach to learn first-order weighted MAX-SAT.
First, clause learning is used to find soft constraints in the given examples.
Then, user-provided preference information is used to learn weights for these constraints.

\subsection{Input} \sam{Dissolve, partially or entirely}
The input consists of global definitions that specify the types and predicates that are used to describe examples and a set of examples \sym{E}.
\anton{Explain: what is type information?}
Type information relates directly to the problem domain and is usually easy to provide.
Including types also improves the accuracy and efficiency of the learned clauses.

Examples are interpretations, they specify the constants in their domain and all predicates that are true.
Predicates can also be generated by background knowledge, in which case they are not included in the example.

When learning optimization criteria, preference information is also provided by the user.
Each preference is of the form $e_1 \succ e_2 \succ ... \succ e_n$ with $e_i \in \sym{E}$.
Preferences describe the relative position of a few examples at a time and are typically easier to provide than absolute scores for all examples.

\begin{example} \sam{Rewrite}
  \label{ex:moving-input}
  Consider the moving problem (example~\ref{ex:moving-input}).
  A model could, for example, use a type \textit{Area} and predicate definitions low\_crime(\textit{Area}), cheap(\textit{Area}), school\_in(\textit{Area}), work\_in(\textit{Area}) and live\_in(\textit{Area}).
  Examples would then specify various areas $A_i$ and describe the conditions (e.g. cheap($A_1$), low\_crime($A_2$)) as well as the users choices (e.g. school\_in($A_2$), work\_in($A_2$), live\_in($A_1$)).
  % Imagine a problem that describes humans with a type \textit{human} and predicates human(\textit{human}), male(\textit{human}) and female(\textit{human}).
  % An example could contain the constants S and A, both of type \textit{human}, and the relations: human(S), human(A), male(S), female(A).
  % Clauses to be learned would be: human(x) $\leftarrow$ male(x), human(x) $\leftarrow$ female(x),  female(x) $\lor$ male(x) $\leftarrow$ human(x) and \textit{false} $\leftarrow$ female(x) $\land$ male(x).
\end{example}

@@

  \begin{algorithmic}
  \State Given: Examples \sym{E} and threshold $t$
  \State $\sym{Q} \gets \{\square\}$, $\sym{T} \gets \{\}$
  \While{$\#\sym{Q} > 0$}
    \State $c \gets next(\sym{Q})$
    \If{$\#\{e \in \sym{E} | e \models c\} \geq t$}
      \If{$\lnot (\sym{T} \models c)$}
        \State $\sym{T} = \sym{T} \cup c$
      \EndIf
    \Else
      \State $\sym{Q} \gets \sym{Q} \cup \rho(c)$
    \EndIf
  \EndWhile
  \State $\sym{T} \gets prune(\sym{T})$
  \State \Return \sym{T}
  \end{algorithmic}
\end{algorithm}

\subsection{Clausal Discovery}
The clause learning system is based on clausal discovery \cite{DeRaedt:ClausalDiscovery}.
Algorithm~\ref{alg:cd} illustrates how clausal discovery works on a high level.
Note the double use of the $\models$ operator.\sam{Review this once yet}
On the one hand, it is used in $e \models c$ to denote that clause~$c$ is true in the example (interpretation)~$e$ and $c$ is said to cover $e$.
This test is used to compute if a clause covers enough examples.
On the other hand, in $\sym{T} \models c$ it signifies that \sym{T} logically entails $c$.
The algorithm uses this test to remove redundant clauses.
For both cases the IDP knowledge base system \cite{de2013prototype,wittocx2008idp} is used to compute $\models$.
The algorithm can find both soft constraints ($t < \# \sym{E}$) or hard constraints ($t = \# \sym{E}$).

Starting with the empty clause ($\square$), clauses are refined until they cover enough examples.
If a clause covers enough examples it is added to the result set, provided it is not entailed by other clauses already in the result set.
In the refinement step, a clause $c = \{l_1, l_2, ..., l_k\}$ is extended (refined) by adding a new literal $l$.
The new clause, $c' = c \cup \{l\}$ is more general than $c$ ($c' \models c$) and potentially covers more examples.
By adding just one literal the refinement operator $\rho$ generates the set of clauses that are minimally more general.
In order to do this efficiently, a list of atoms is pre-calculated, given a maximal number of variables that can occur within a clause.
Within the body and head of a clause, atoms may only be added in a specific order to avoid generating redundant clauses.

A few additional measures are used to enable the implementation to be more efficient.
Most importantly, there are two syntactic restrictions on the form of the clauses.
Clauses must be connected (i.e. new atoms must always contain a variable that has already occurred) and range-restricted (i.e. no new variables may be introduced in the head of the clause).
Moreover, Object Identity is used to specify that variables with different names must denote different objects.
Therefore, a clause~$c$ with variables \sym{V} is extended to the clause~$c_{OI}$ by adding literals $x_i \neq x_j$, for $x_i, x_j \in \sym{V}, i \neq j$.
Finally, the implementation natively supports symmetric predicates for which the order of the terms do not matter

\subsection{Optimization}
In order to find weighted soft constraints, soft constraints~\sym{C} for the given examples are learned by using the constraint learning system with a (low) threshold.
The example are then translated to Boolean feature vectors of length $\# \sym{C}$ by introducing a feature for every clause~$c_i \in \sym{C}$.
For an example~$e$, the $i$-th value of its feature vector $\mathbf{f}_e$ is calculated as $\mathbf{f}_e(i) = \mathbf{1}_{e \models c_i}$.
Next, existing software is used to learn the linear function $\sum_i w_i \cdot c_i$ over these features, based on user provided preferences.
This function corresponds with the scoring function (\ref{eqn:score})) introduced in Section~\ref{sec:problem}.
It can be used to calculate scores for (unseen) examples and by maximizing the score function an optimal solution can be found.

\begin{example} \sam{Rewrite}
  \label{ex:features}
  Consider examples~$e_1, e_2, e_3$, rankings $e_1 \succ e_2, e_2 \succ e_3$ and clauses $c_1, c_2, c_3$.
  If the clauses cover examples according to table~\ref{tbl:cover_examples}, then, for example, the function $(1 \cdot c_1) + (0\cdot c_2) + (-2\cdot c_3)$ perfectly models the given rankings.
  A new example that is covered by~$c_1$ and~$c_2$, would be assigned a score of $1$, according to this function.
  This score has no value in the absolute sense, it can only be used to compare it to other examples ranked by the same function.
  In this example clause~$c_1$ represents a desirable property, clause~$c_2$ is ignored because it does not influence the ranking and clause~$c_3$ represents an undesirable property.

  \begin{table}
  \caption{Clause coverage}
  \label{tbl:cover_examples}
  \begin{tabularx}{\linewidth}{c|l|X}
    \textbf{Example} & \textbf{Covered by} & \textbf{Feature vector}\\
    \toprule
    $e_1$             & $c_1$               & $\mathbf{f}_{e_1} = (1, 0, 0)$ \\
    $e_2$             & $c_2$               & $\mathbf{f}_{e_2} = (0, 1, 0)$ \\
    $e_3$             & $c_1$, $c_2$, $c_3$ & $\mathbf{f}_{e_3} = (1, 1, 1)$ \\
  \end{tabularx}
  \end{table}
\end{example}

\svm{} \cite{joachims2006training} was chosen to find the scoring function since it uses a linear model and offers a robust and efficient implementation.
The weights that it assigns to the features can be used directly as weights for the optimization criteria.
While this paper uses \svm{} other linear ranking systems such as Coordinate Ascent \cite{metzler2007linear} could also be used.

\subsection{Optimal Solution}
Solvers, such as IDP \sam{more on IDP, can be used as ASP solver?}, can use clauses directly as hard constraints to generate a solution.
There is no solver that can use the chosen optimization criteria directly.
Weighted MAX-SAT solvers use propositional clauses and only allow for positive weights.

This optimization task can be solved in IDP, using inductive definitions, aggregates and minimization.
The only limitation is that the current version only supports integer values.
Therefore, the weights of the clauses are divided by the smallest absolute weight, multiplied by a constant and rounded to the closest integer if necessary.

@@

\section{Evaluation}
In our experiments we aim to measure the accuracy and efficiency of the learning systems for constraints and optimization criteria.
In order to account for non-determinism, experiments measuring execution times or optimization scores were repeated eight times and their result were averaged.
\sam{First step evaluation constraint learning, then optimization}
\subsection{Constraints}
We used four problems to evaluate constraint learning: \sam{clarify hard constraints}
\begin{description}
\item[Map Coloring] Countries are assigned colors and neighboring countries may not have the same color.
Two examples, each containing three correctly colored countries, are given.
\item [Sudoku] A single, solved $4 \times 4$ sudoku is given.
\item[Elevator] Three examples have been generated, two of which respect the underlying soft constraint.
\item[Co-housing] Contains four hard constraints and five examples have been generated that respect the constraints.
\end{description}
\anton{more elaborate descriptions of these?}
\sam{link to thesis}

\begin{question}[Accuracy]
  Are the essential constraints discovered and what influence do different parameters have on the accuracy?
\end{question}
For all problems the essential constraints are found.
Often additional constraints were found that describe some structure in the problems.
For example, for the map coloring problem one of the learned constraint states that countries are never their own neighbor.
These kind of constraints may help a constraint solver to work more efficiently.

@@

Table~\ref{tbl:uitvoering} shows the execution times for the different problems.
It shows that smaller problems can be solved efficiently.

@@

\section{Conclusion}
We presented an approach to automatically acquire constraints and optimization criteria from examples and preferences.
Implementation for both clause learning and clausal optimization have been provided that accomplish these tasks using first-order logic clauses.

>>>>

\graphicspath{{Graphics/}{}}

@@

\newtheorem{question}{Question}
\theoremstyle{definition}
\newtheorem{example}{Example}

\newcommand{\sam}[1]{{\color{red}(#1)}}
\newcommand{\abs}[1]{\left|#1\right|}

@@

\begin{abstract}
While there exist several approaches in the constraint programming community to learn a constraint theory, few of them have considered the learning of constraint optimization problems.
To alleviate this situation, we introduce an initial approach to learning first-order weighted MAX-SAT theories. 
It employs inductive logic programming techniques to learn a set of first-order clauses and then uses preference learning techniques to learn the weights of the clauses.
In order to learn these weighted clauses, the clausal optimization system uses examples of possible worlds and a set of preferences that state which examples are preferred over other ones.
The technique is also empirically evaluated on a number of examples.
These experiments show that the system is capable of learning clauses and weights that accurately capture underlying models.
\end{abstract}

@@

\section{Introduction}
\sam{it might be good to state somewhere early that you are in the "learning from interpretations" setting, but additionally have preferences}
\sam{Structure}
An important class of declarative models is concerned with (discrete) constrained optimization problems.
These combine constraint satisfaction problems with an optimization function that specifies which solutions are optimal.
While these models are powerful and elegant, they are often also hard to obtain, certainly for non-experts \cite{Wallace:PrinciplesCP}.
This has motivated several researchers to investigate whether such models can be learned.

Learning constraints has been addressed in constraint programming (e.g. \cite{Beldiceanu:ModelSeeker,bessiere2013constraint}) and inductive logic programming (e.g. \cite{DeRaedt:ClausalDiscovery}).
Recent efforts \cite{Lallouet:LearningCP} have attempted to combine these two fields.
Furthermore, \cite{campigotto2011active} have learned propositional weighted MAX-SAT theories from examples.
Also, work in statistical relational learning (SRL) on learning the structure of Markov Logic networks (MLN) or other SRL models \cite{kok2005learning} is relevant, as MAP inference can be used to query for the most likely state.

@@

\paragraph{Clauses}
We use (function free) first-order clauses; which we now introduce.
An atom $p(t_1, t_2, ..., t_n)$ consists of a predicate symbol~$p$ that is applied to $n$ terms~$t_i$. \sam{Casing}
Terms are either constants or variables.
In this paper constants will be written in uppercase ($\mathit{John}$) and variables in lowercase ($\mathit{x}$).
A literal can be an atom~$a$ (positive literal) or a negated atom~$\lnot a$ (negative literal).
Clauses are disjunctions of literals $l_1 \lor l_2 \lor ... \lor l_k$ and are assumed to be universally quantified.
A clause can be expressed using a body and a head $\mathit{head} \leftarrow \mathit{body}$.
Hereby, positive literals are grouped in the head and negative literals are grouped in the body.
Sometimes clauses are also written as sets of literals $\{l_1, l_2, ..., l_k\}$.

@@

\paragraph{Optimization}
The Boolean Satisfiability problem (SAT) is the problem of determining whether a propositional formula can be satisfied.
Formulas can be rewritten into Conjunctive Normal Form (CNF), which consists of a conjunction of clauses (disjunctions).
An extension of this problem is the Maximum Satisfiability Problem (MAX-SAT), which is an optimization problem and attempts to maximize the number of satisfied clauses.
By assigning positive integer weights to clauses and maximizing the sum of the weights of satisfied clauses, one obtains the weighted MAX-SAT problem.

\sam{versus MLN, PSL}
In related work \cite{campigotto2011active} weighted MAX-SAT problems are induced by learning propositional clauses and weights using examples and absolute scores.
The representation we use is an extension of weighted MAX-SAT to first-order logic.
We also learn this model using preferences over sets of examples (possible worlds), rather than assigning target values to examples.

\paragraph{Constraint Learning}
\sam{Define constraints}
The goal of constraint learning is to automatically identify constraints in one or multiple examples.
This task is difficult because the search space of constraints is typically very large and, unlike in typical machine learning problems, there are usually only few examples to learn from.

@@

\section{Problem Statement}
\label{sec:problem}
We want to learn constraints and optimization criteria based on positive examples and user preferences.
Since our aim is to simplify the process of generating formal representations, any input should be reasonably easy for a user to provide.

In order to apply ILP techniques for constraint learning, we will use first-order logical clauses as constraints.

\paragraph*{Clause Learning}
\emph{Given a set of examples and a threshold $t$, find maximally specific clauses that are satisfied by at least $t$ of the given examples.}
\\\\
Examples are Herbrand interpretations.
They consist of a set of constants, the domain, and all ground atoms over these constants that are true.
The learned clauses are domain-independent hard or soft constraints and contain only variables.
\\\\
Our goal is to learn optimization criteria, weighted soft constraints, based on examples (interpretations) and user preferences.
A (pairwise) preference $e_1 \succ e_2$ describes a relative order over examples $e_1$ and~$e_2$, stating that $e_1$ is preferred over~$e_2$.
Preferences only consider two examples at a time and are usually easier to provide than absolute scores.
The soft constraints consist of clauses~(\sym{C}) that are satisfied by some but not all of the examples.
They can be identified using a system that implements the constraint learning task.
Weighted clauses~($\sym{W}_{\sym{C}}$) are tuples $\mathbb{R} \times \sym{C}$ that can be used to calculate a score for any example $e$:
\begin{eqnarray}
  \label{eqn:score}
  score_{\sym{W}_{\sym{C}}}(e) = \sum\limits_{(\mathit{w}, \mathit{c}) \in \sym{W}_{\sym{C}}} \mathit{w} \cdot \mathbf{1}_{e \models c}
\end{eqnarray}
An interpretation~$e_{opt}$ is an optimal solution with respect to~$\sym{W}_{\sym{C}}$ if $score_{\sym{W}_{\sym{C}}}(e_{opt})$ is maximal.

@@

  Assume the following set of weighted clauses~\sym{M} describes the optimization problem to solve:
  \begin{flalign*}
    (w_1, c_1) &= (0.50, \mathit{low\_crime}(a) \leftarrow \mathit{live\_in}(a)) \\
    (w_2, c_2) &= (0.25, \mathit{school\_in}(a) \leftarrow \mathit{work\_in}(a)) \\
    (w_3, c_3) &= (1.00, \mathit{low\_crime}(a) \leftarrow \mathit{school\_in}(a)) \\
    (w_4, c_4) &= (-1.00, \mathit{false} \leftarrow \mathit{live\_in}(a) \land \mathit{cheap}(a))
  \end{flalign*}
  Let examples $e_1$, $e_2$ and~$e_3$ be Herbrand interpretations that reason over the same city, using the same constants (areas) $A_1$, $A_2$ and $A_3$.
  Consider the set~\sym{S} of shared ground atoms $\{\mathit{cheap}(A_1)$, $\mathit{cheap}(A_3)$, $\mathit{low\_crime}(A_1)$, $\mathit{low\_crime}(A_2)\}$ and define the examples as:
  \begin{flalign*}
    &e_1 = \{\mathit{live\_in}(A_1), \mathit{work\_in}(A_2), \mathit{school\_in(A_2)}\} \cup \sym{S} \\
    &e_2 = \{\mathit{live\_in}(A_2), \mathit{work\_in}(A_2), \mathit{school\_in(A_3)}\} \cup \sym{S}\\
    &e_3 = \{\mathit{live\_in}(A_2), \mathit{work\_in}(A_2), \mathit{school\_in(A_1)}\} \cup \sym{S}.
  \end{flalign*}

  The score of an example can be calculated using~(\ref{eqn:score}).
  Example~$e_1$ is covered by clauses $c_1$, $c_2$ and~$c_3$, therefore, $score_{\sym{M}}(e_1) = 0.5 + 0.25 + 1 + 0 = 1.75$.
  Analogously we can compute $score_{\sym{M}}(e_2) = -1$ and $score_{\sym{M}}(e_3) = 0$.
  Using $score_{\sym{M}}$, the following preferences can be generated: $e_1 \succ e_2$, $e_1 \succ e_3$ and $e_2 \succ e_3$.
\end{example}

The aim of our research is to learn weighted clauses such that the order imposed by these optimization criteria maximally corresponds with the given preference information.

\paragraph*{Clausal Optimization}
\emph{Given examples~\sym{E}, preferences~\sym{P} and threshold~$t$, learn clauses~\sym{C} that cover at least $t$ examples and find weights~$\sym{W}_{\sym{C}}$ that maximize:
\begin{flalign*}\abs{\{e_1 \succ e_2 \in \sym{P} | score_{\sym{W}_{\sym{C}}}(e_1) > score_{\sym{W}_{\sym{C}}(e_2, })\}}.\end{flalign*}}

@@

\section{Approach}
We use a two-step approach to learn weighted first-order clauses.
First, clause learning is used to find soft constraints in the given examples.
Then, user-provided preference information is used to learn weights for these constraints.

% \begin{example}
%   \label{ex:moving-input}
%   Consider the moving problem (example~\ref{ex:moving-input}).
%   A model could, for example, use a type \textit{Area} and predicate definitions low\_crime(\textit{Area}), cheap(\textit{Area}), school\_in(\textit{Area}), work\_in(\textit{Area}) and live\_in(\textit{Area}).
%   Examples would then specify various areas $A_i$ and describe the conditions (e.g. cheap($A_1$), low\_crime($A_2$)) as well as the users choices (e.g. school\_in($A_2$), work\_in($A_2$), live\_in($A_1$)).
%   % Imagine a problem that describes humans with a type \textit{human} and predicates human(\textit{human}), male(\textit{human}) and female(\textit{human}).
%   % An example could contain the constants S and A, both of type \textit{human}, and the relations: human(S), human(A), male(S), female(A).
%   % Clauses to be learned would be: human(x) $\leftarrow$ male(x), human(x) $\leftarrow$ female(x),  female(x) $\lor$ male(x) $\leftarrow$ human(x) and \textit{false} $\leftarrow$ female(x) $\land$ male(x).
% \end{example}

@@

  \begin{algorithmic}
  \State Given: Examples \sym{E} and threshold $t$
  \State $\sym{Q} \gets \{\square\}$, $\sym{T} \gets \{\}$
  \While{$\abs{\sym{Q}} > 0$}
    \State $c \gets next(\sym{Q})$
    \If{$\abs{\{e \in \sym{E} | e \models c\}} \geq t$}
      \If{$\sym{T} \not\models c$}
        \State $\sym{T} = \sym{T} \cup c$
      \EndIf
    \Else
      \State $\sym{Q} \gets \sym{Q} \cup \rho(c)$
    \EndIf
  \EndWhile
  \State $\sym{T} \gets prune(\sym{T})$
  \State \Return \sym{T}
  \end{algorithmic}
\end{algorithm}

\subsection{Clause Learning}
The clause learning system is based on clausal discovery \cite{DeRaedt:ClausalDiscovery}.
Algorithm~\ref{alg:cd} illustrates how clausal discovery works on a high level.
To compute if a clause~$c$ covers enough examples, the number of examples~$e$ is counted for which $e \models c$ (i.e. $c$ is true in the example~$e$ or $c$ covers $e$).
In order to remove redundant clauses a new clause~$c$ is only added to the result set $\sym{T}$ if $\sym{T} \not\models c$ (i.e. $c$ is not logically entailed by \sym{T}).
Both $e \models c$ and $\sym{T} \not\models c$ are computed by the IDP knowledge base system \cite{de2013prototype,wittocx2008idp}.
The algorithm can find both soft constraints ($t < \abs{\sym{E}}$) or hard constraints ($t = \abs{\sym{E}}$).

Starting with the empty clause ($\square$), clauses are refined until they cover enough examples.
If a clause covers enough examples it is added to the result set, provided it is not entailed by other clauses already in the result set.
In the refinement step, a clause $c = \{l_1, l_2, ..., l_k\}$ is extended (refined) by adding a new literal $l$.
The new clause, $c' = c \cup \{l\}$ is more general than $c$ ($c' \models c$) and potentially covers more examples.
By adding just one literal the refinement operator $\rho$ generates the set of clauses that are minimally more general.
In order to do this efficiently, our implementation uses a pre-calculated list of atoms, given a maximal number of variables that can occur within a clause.
Within the body and head of a clause, atoms may only be added in a specific order to avoid generating redundant clauses.

A few additional measures are used to make the implementation more efficient.
Most importantly, there are two syntactic restrictions on the form of the clauses.
Clauses must be connected (i.e. new atoms must always contain a variable that has already occurred) and range-restricted (i.e. no new variables may be introduced in the head of the clause).
Moreover, Object Identity is used to specify that variables with different names must denote different objects.
Therefore, a clause~$c$ with variables \sym{V} is extended to the clause~$c_{OI}$ by adding literals $x_i \neq x_j$, for $x_i, x_j \in \sym{V}, i \neq j$.
Finally, the implementation natively supports typed constants and predicates, symmetric predicates (i.e. predicates where the order of terms does not matter) and background knowledge.
A predicate can be marked as \emph{calculated} in which case it is defined in the background knowledge, rather than being provided in the examples.

\subsection{Clausal Optimization}
In order to find weighted soft constraints, the clause learning system is used with a low threshold to learn (soft) clauses~\sym{C}.
The examples are then translated to Boolean feature vectors of length $\abs{\sym{C}}$ by introducing a feature for every clause~$c_i \in \sym{C}$.
For an example~$e$, the value of its feature vector $\mathbf{v}_e$ for clause~$c_i$ is calculated as $\mathbf{v}_e(i) = \mathbf{1}_{e \models c_i}$.
Next, a linear function $f(\mathbf{v})=\sum_i w_i \cdot \mathbf{v}(i)$ is learned that assigns weights~$w_i$ to clauses~$c_i$, based on the user-provided preferences.
The scoring function~$f$ agrees with a preference $e_1 \succ e_2$ if $f(\mathbf{v}_{e_1}) > f(\mathbf{v}_{e_2})$ and corresponds to the scoring function (\ref{eqn:score}).

\begin{example}
  \label{ex:features}
  Consider examples~$e_1, e_2, e_3$, preferences $e_1 \succ e_2, e_2 \succ e_3$ and clauses $c_1, c_2, c_3$, whereby clauses cover examples according to Table~\ref{tbl:cover_examples}.
  The function $f(\mathbf{v}) = (1, 0, -2) \cdot \mathbf{v}$ assigns scores $f(\mathbf{v_{e_1}}) = 1$, $f(\mathbf{v_{e_2}}) = 0$ and $f(\mathbf{v_{e_3}}) = -1$ and perfectly models the given preferences since $f(\mathbf{v_{e_1}}) > f(\mathbf{v_{e_2}})$ and $f(\mathbf{v_{e_2}}) > f(\mathbf{v_{e_3}})$.
  This score has no value in the absolute sense, it can only be used to compare between examples.
  In this example clause~$c_1$ represents a desirable property and clause~$c_3$ represents an undesirable property.

  \begin{table}
  \caption{Clause coverage}
  \label{tbl:cover_examples}
  \begin{tabularx}{\linewidth}{c|l|X}
    \textbf{Example} & \textbf{Covered by} & \textbf{Feature vector}\\
    \toprule
    $e_1$             & $c_1$               & $\mathbf{v}_{e_1} = (1, 0, 0)$ \\
    $e_2$             & $c_2$               & $\mathbf{v}_{e_2} = (0, 1, 0)$ \\
    $e_3$             & $c_1$, $c_2$, $c_3$ & $\mathbf{v}_{e_3} = (1, 1, 1)$ \\
  \end{tabularx}
  \end{table}
\end{example}

\svm{} \cite{joachims2006training} was chosen to find the scoring function since it uses a linear model and offers a robust and efficient implementation.
The weights that it assigns to the features can be used directly as weights for the clauses.
While the current implementation uses \svm{}, other linear ranking systems such as Coordinate Ascent \cite{metzler2007linear} could also be used.

\subsection{Optimal Solution}
Typical MAX-SAT solvers use propositional clauses and only allow for positive weights.
However, the first-order logic solver IDP can solve our optimization task as it supports inductive definitions, aggregates and minimization.
The only limitation is that the current version is restricted to integer values and does not support rational numbers.
Therefore, the weights of the clauses are divided by the smallest absolute weight, multiplied by a constant and rounded to the closest integer.

@@

\section{Evaluation}
In our experiments we aim to measure the accuracy and efficiency of the learning systems for constraints and optimization criteria.
Non-deterministic experiments (e.g. measuring execution times or accuracy of clausal optimization) were repeated eight times and their result were averaged.
Some problems and experiments are only described briefly or summarized, for more details we refer to \cite{kolb2015thesis}\footnote{Available online at https://github.com/samuelkolb/thesis}.

First, experiments concerning the accuracy and efficiency of the constraint learning implementation are discussed.
Thereafter, we discuss the evaluation of the accuracy of the clausal optimization system\footnote{The system (v1.0) used to perform all experiments is available online at https://github.com/samuelkolb/clausal-discovery/releases}.

\subsection{Clause Learning}
We used four problems to evaluate constraint learning:
\begin{description}
\item[Map Coloring] Countries are assigned colors and neighboring countries may not have the same color.
Two examples, each containing three correctly colored countries, are given.
\item[Sudoku] A single, solved $4 \times 4$ sudoku is given as an example for this problem.
\item[Elevator] This problem uses three predicates $\mathit{inside}$, $\mathit{crowded}$ and $\mathit{panic}$, with $\mathit{crowded}$ being a calculated predicate.
It is defined by background knowledge to be true for all elevators with three or more people inside.
Out of the three examples provided for this problem, two satisfy the target soft constraint:
\begin{flalign*}
  &\mathit{panic}(p)~\leftarrow~\mathit{crowded}(e)~\land~\mathit{inside(p, e)}.
\end{flalign*}
\item[Co-housing] The co-housing problem also uses background knowledge to specify one of its predicates ($\mathit{cheap}$) and to impose some additional constraints.
Five examples are provided, which all adhere to the following four constraints on the housing choices of two friends:
\begin{enumerate}
  \item If the two friends do not live in the same area, they work in the same area.
  \item If a person does not work and live in the same area, that person needs a car.
  \item A person who lives in a cheap area has a car.
  \item If the two friends live together, they live in an expensive area.
\end{enumerate}
\end{description}
The clausal discovery implementation is used to learn soft constraints for the elevator problem ($t = 1$) and hard constraints for the other problems.

\begin{question}[Accuracy]
  Are the essential constraints discovered and what influence do different parameters have on the accuracy?
\end{question}
For all problems the essential constraints are found.
Often additional constraints were found that describe some structure in the problems.
For example, for the map coloring problem one of the learned constraints states that countries are never their own neighbor.
These kind of constraints may help a constraint solver to work more efficiently.

@@

Table~\ref{tbl:uitvoering} shows the execution times for the different problems\footnote{All experiments were run on a MacBook Pro with an i7 quad-core processor}.
It shows that the evaluated problems can be solved efficiently.

@@

\section{Conclusion}
We presented an approach to automatically acquire constraints and optimization criteria from examples and preferences.
Implementations for both clause discovery and clausal optimization have been provided that accomplish these tasks using first-order logic clauses.
